---
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```



# Yield Curve inversion

Every so often, we hear warnings from commentators on the "inverted yield curve" and its predictive power with respect to recessions. An explainer what a [inverted yield curve is can be found here](https://www.reuters.com/article/us-usa-economy-yieldcurve-explainer/explainer-what-is-an-inverted-yield-curve-idUSKBN1O50GA).

_But are inversions are truly harbingers of recessions?_


To analyse this, I will look at US data and use the [FRED database](https://fred.stlouisfed.org/) to download historical yield curve rates, and plot the yield curves since 1999 to see when the yield curves flatten.


First, use the `tidyquant` package to download monthly rates for different durations. 

```{r get_rates, warning=FALSE}
# Get a list of FRED codes for US rates and US yield curve; choose monthly frequency
# to see, eg., the 3-month T-bill https://fred.stlouisfed.org/series/TB3MS
tickers <- c('TB3MS', # 3-month Treasury bill (or T-bill)
             'TB6MS', # 6-month
             'GS1',   # 1-year
             'GS2',   # 2-year, etc....
             'GS3',
             'GS5',
             'GS7',
             'GS10',
             'GS20',
             'GS30')  #.... all the way to the 30-year rate

# Turn  FRED codes to human readable variables
myvars <- c('3-Month Treasury Bill',
            '6-Month Treasury Bill',
            '1-Year Treasury Rate',
            '2-Year Treasury Rate',
            '3-Year Treasury Rate',
            '5-Year Treasury Rate',
            '7-Year Treasury Rate',
            '10-Year Treasury Rate',
            '20-Year Treasury Rate',
            '30-Year Treasury Rate')

maturity <- c('3m', '6m', '1y', '2y','3y','5y','7y','10y','20y','30y')

# by default R will sort these maturities alphabetically; but since we want
# to keep them in that exact order, we recast maturity as a factor 
# or categorical variable, with the levels defined as we want
maturity <- factor(maturity, levels = maturity)

# Create a lookup dataset
mylookup<-data.frame(symbol=tickers,var=myvars, maturity=maturity)
# Take a look:
mylookup %>% 
  knitr::kable()

df <- tickers %>% tidyquant::tq_get(get="economic.data", 
                   from="1960-01-01")   # start from January 1960

glimpse(df)
```

The dataframe `df` has three columns (variables):

- `symbol`: the FRED database ticker symbol
- `date`: already a date object
- `price`: the actual yield on that date

The first thing would be to join this dataframe `df` with the dataframe `mylookup` so we have a more readable version of maturities, durations, etc.


```{r join_data, warning=FALSE}

yield_curve <-left_join(df,mylookup,by="symbol") 
```



### Yields on US rates by duration since 1960

```{r yield_curve_1A, out.width="100%"}

names(myvars) <- maturity #links myvars with maturity
yield_curve %>%
  group_by(maturity) %>%
  ggplot(aes(x = date, y = price, colour = maturity))+
  geom_line() +
  facet_wrap(~maturity, ncol = 2, labeller = labeller(maturity = myvars)) + #labeller ensures the facet labels are taken from myvars
  theme(legend.position = "none") + #removes legend 
  labs(title = "Yields on US rates since 1960", caption = "Source: St. Louis Federal Reserve Economic Database (FRED)") + 
  xlab("") + 
  ylab("Yield(%)")


```



### Monthly yields on US rates by duration since 1999 on a year-by-year basis

```{r yield_curve_2A, out.width="100%"}
yield_curve_99 <- yield_curve %>%
  #commented Mutate() below is another way ofpulling Year/Month from dates
  #mutate(year = format(as.Date(yield_curve$date, format = "%Y/%m/%d"), "%Y"))
  mutate(year = year(lubridate::ymd(yield_curve$date)), month = month(lubridate::ymd(yield_curve$date))) %>% #adds Year & month column determined from date column
  filter(year >= 1999) %>%
  group_by(year)

ggplot(yield_curve_99, aes(x = maturity, y = price), colour = year)+
  geom_line(aes(group=month, colour = year)) +
  facet_wrap(~year, ncol = 4) + 
  theme(legend.position = "none") + 
   labs(title = "US Yield Curve", caption = "Source: St. Louis Federal Reserve Economic Database (FRED)") + 
  xlab("Maturity") + 
  ylab("Yield(%)")
```


### 3-month and 10-year yields since 1999

```{r yield_curve_3, out.width="100%"}
#filter out only 3m Rates & 10y Bills
yield_curve_99 %>%
  ungroup() %>%
  filter (maturity == '3m' | maturity == '10y' ) %>%
  ggplot(aes(x = date, y = price, colour = maturity))+
  geom_line() +
   labs(title = "Yields on 3-month and 10-year US Treasury rates since 1999", caption = "Source: St. Louis Federal Reserve Economic Database (FRED)") + 
  xlab("") + 
  ylab("Yield(%)")
```


According to [Wikipedia's list of recession in the United States](https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States), since 1999 there have been two recession in the US: between Mar 2001–Nov 2001 and between Dec 2007–June 2009.

Here are questions to consider:
- Does the yield curve seem to flatten before these recessions? 
- Can a yield curve flattening really mean a recession is coming in the US?
- Since 1999, when did short-term (3 months) yield more than longer term (10 years) debt?



Before calculating the spread (10year - 3months), there are a few things to do to produce our final plot

1. Setup data for US recessions 
1. Superimpose recessions as the grey areas in our plot
1. Plot the spread between 30 years and 3 months as a blue/red ribbon, based on whether the spread is positive (blue) or negative(red)


- For the first, the code below creates a dataframe with all US recessions since 1946

```{r setup_US-recessions, warning=FALSE}

# get US recession dates after 1946 from Wikipedia 
# https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States

recessions <- tibble(
  from = c("1948-11-01", "1953-07-01", "1957-08-01", "1960-04-01", "1969-12-01", "1973-11-01", "1980-01-01","1981-07-01", "1990-07-01", "2001-03-01", "2007-12-01"),  
  to = c("1949-10-01", "1954-05-01", "1958-04-01", "1961-02-01", "1970-11-01", "1975-03-01", "1980-07-01", "1982-11-01", "1991-03-01", "2001-11-01", "2009-06-01") 
  )  %>% 
  mutate(From = ymd(from), 
         To=ymd(to),
         duration_days = To-From)

recessions
```


### Evaluating US Yield Curve as a predictor of recessions
```{r}
#remove from/to columns without date format
recessions_clean <- recessions %>%
  select(From, To, duration_days)

#filters relevant maturities and pivots wider
yield_spread <- yield_curve %>%
  filter (maturity == '3m' | maturity == '10y'| maturity == '30y') %>%
  select(date, maturity, price) %>%
  pivot_wider(names_from = maturity, values_from = price)

#Adds 10y-3m spread and 30y-3m spread
yield_spread <- yield_spread %>%
  mutate (spread_10 = yield_spread$'10y' - yield_spread$'3m') %>% 
  mutate (spread_30 = yield_spread$'30y' - yield_spread$'3m')



ggplot(data=yield_spread,aes(x=date,y=spread_10))+
  
  #geom_rect below plots grey bars fo recessions
  geom_rect(data=filter(recessions_clean,year(From)>=1960), inherit.aes=F, aes(xmin=From, xmax=To, ymin=-Inf, ymax=+Inf), fill='darkgray', alpha=0.5) +
  
  geom_line(color="black")+
  
  theme_minimal()+
  
  #geom_ribbon colours negative spread red
  geom_ribbon(aes(ymin=0,ymax=ifelse(spread_10 < 0, spread_10, 0)),fill="#d73027",alpha=0.5)+
  
  #geom_ribbon colours positive spread blue
  geom_ribbon(aes(ymin=ifelse(spread_10 > 0, spread_10, 0),ymax=0),fill="#4575b4",alpha=0.5)  +
  
  #geom_rug adds ribbon at bottom to indicate positive/negative spreads
  geom_rug(aes(color=ifelse(spread_10>=0,">=0","<0")),sides="b") +
  
  theme(legend.position = "none") + 
  
  labs(title = "Yield Curve Inversion: 10-year minus 3-month US Treasury Rates", caption = "Source: St. Louis Federal Reserve Economic Database (FRED)") + 
  xlab("") + 
  ylab("Yield(%) Spread Betwwen 10-year and 3-month rates")


```

